###############################################################################
# Project:    ANNOTATION AND PHYLOGENETIC ANALYSIS OF SACCHAROMYCES GENES     #
#             RELATED WITH CARBON SOURCES OF BREWING INTEREST                 #
# Author:     Laura Varón Pozuelo                                             #
# Date:       2024-10-27                                                      #  
# Script:     analysis_pipeline.Rmd (RMarkdown)                               #  
#                                                                             #
# Description:                                                                #
# This pipeline processes proteome datasets to:                               #
#   1. Import reference genes from FASTA.                                     #
#   2. Parse proteome FASTA files per strain (contigs).                       # 
#   3. Load BLAST results against target genes.                               #          
#   4. Select best hits (filters + overlaps).                                 #
#   5. Annotate strain species.                                               #  
#   6. Identify copy number variation (CNV).                                  #
#   8. Export curated sequences per strain in FASTA format.                   #  
#                                                                             #
# Input files/directories:                                                    #  
#   - general/genes_aa.fasta        (reference protein sequences)             #
#   - proteomes/*.fasta|fa          (strain proteomes)                        #
#   - blast/*.tsv                   (BLAST results)                           #   
#   - proteomes/aa_hits/*.faa (amino acid hits)                               #
#                                                                             #
# Outputs:                                                                    #
#   - CNV.tsv                       (copy number variation summary)           #
#   - proteomes/aa_hits/*.faa (exported FASTA per strain)                     #
#                                                                             #
###############################################################################


```{r setup, include=FALSE}
# Project setup
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
rm(list=ls())
library(pacman)
pacman::p_load(
  dplyr, tidyr, ggplot2, ggtext, viridis, stringr, purrr, Biostrings)
base_dir      <- "."
proteomes_dir <- file.path(base_dir, "proteomes")
blast_dir     <- file.path(base_dir, "blast")
genes_dir     <- file.path(base_dir, "general")
stats_dir     <- file.path(base_dir, "stats")
hits_dir      <- file.path(base_dir, "hits")
faa_dir       <- file.path(proteomes_dir, "aa_hits")
```


```{r helpers}
# Helper functions

# Keep only the target genes of interest (IMA, MAL, MTT1/MTY1 families)
valid_genes <- function(df) {
  df %>% filter(
    grepl("^IMA[0-3]$", gene) |
    gene == "IMA5" |
    grepl("^MAL[0-9]{2}$", gene) |
    grepl("^MT[T|Y]1$", gene)
  )
}

# Check if two intervals [S1,E1] and [S2,E2] overlap
overlap <- function(S1, E1, S2, E2) {
  !(E1 < S2 || E2 < S1)
}
```


```{r genes}
# ====================
# Import and parse gene fasta file
# ====================

# Read fasta file with gene sequences
genes_lines <- readLines(file.path(genes_dir, "genes_aa.fasta"))

# Extract headers (lines starting with ">")
genes_info  <- gsub("^>", "", grep("^>", genes_lines, value = TRUE))

# Build tibble with basic gene information
genes_df <- tibble(
  gene       = sub(" .*", "", genes_info), # gene name
  reference  = str_extract(
    genes_info, "(?<=organism=)[^]]+"), # reference organism
  chromosome = str_extract(
    genes_info, "(?<=chromosome=)[^]]+") # chromosome info
) %>%
  mutate(across(everything(), ~replace_na(.x, "")))

head(genes_df)
```


```{r contigs}
# ====================
# Parse contig names for each strain
# ====================

# Get all proteome fasta files
strain_files <- list.files(proteomes_dir, pattern = "\.(fasta|fa)$")
strains      <- gsub("\.(fasta|fa)$", "", strain_files)

# Extract contig names from fasta headers
contigs_list <- map(strain_files, function(f) {
  contigs <- grep("^>", readLines(file.path(proteomes_dir, f)), value = TRUE)
  sapply(
    strsplit(
      sub(
        "^>", paste0(gsub("\.(fasta|fa)$", "", f), "_", contigs)), " "), `[`, 1)
})

names(contigs_list) <- strains

# Combine into wide dataframe
contigs_df <- bind_cols(contigs_list)
head(contigs_df)
```


```{r blast}
# ====================
# Read BLAST output results
# ====================

# Load BLAST results (tab-delimited TSV format)
blast_files <- list.files(blast_dir, pattern = "\.tsv$", full.names = TRUE)

df <- map_dfr(blast_files, read.delim, sep = "\t", .id = "source") %>%
  mutate(
    strain = sub("_.*", "", accession),         # extract strain from accession
    chr    = str_extract(accession, "chr[^_\-]*")      # extract chromosome id
  ) %>%
  filter(X.match >= 90, coverage >= 90) # quality thresholds

head(df)

```


```{r correct_hits}
# ====================
# Select the best hit per gene per strain
# ====================

# Split BLAST df by strain
dataframes_list <- split(df, df$strain)

# Process each strain separately
correct_hits <- map(dataframes_list, function(d) {
  if (nrow(d) <= 1) return(valid_genes(d)) # shortcut if only one hit
  
  d <- d %>%
    arrange(accession, start) %>%
    mutate(group = 0)
  
  group_counter <- 0
  for (j in 2:nrow(d)) {
    same_acc <- d$accession[j] == d$accession[j-1]
    if (same_acc && overlap(d$start[j-1], d$end[j-1], d$start[j], d$end[j])) {
      # Group overlapping hits
      d$group[c(j-1, j)] <- group_counter
    } else {
      # Otherwise, start a new group
      group_counter <- group_counter + 1
      d$group[j] <- group_counter
    }
  }
  
  # Within each group keep the best hit (by score and coverage)
  d %>%
    group_by(group) %>%
    slice_max(order_by = identity, n = 1, with_ties = FALSE) %>%
    slice_max(order_by = coverage, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    valid_genes()
})

# Concatenate results across strains
combined_df <- bind_rows(correct_hits, .id = "strain")
head(combined_df)
```


```{r strain_species}
# ====================
# Annotate species for each strain
# ====================

strain_species <- read.delim(
  file.path(genes_dir, "strain_species.tsv"), 
  sep = "\t", header = TRUE, stringsAsFactors = FALSE
) %>%
  as_tibble()

# Add species annotation to combined BLAST results
combined_df <- left_join(combined_df, strain_species, by = "strain")
head(combined_df)
```


```{r cnv}
# ====================
# Detect Copy Number Variation (CNV)
# ====================

# Count gene copies per strain and chromosome
CNV_correct <- combined_df %>%
  filter(!is.na(group)) %>%  
  count(gene, species, strain, accession, name = "n_copies") %>%
  arrange(strain)

# Save CNV results to file
write.table(
  CNV_correct, file = "CNV.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

# Collapse by strain/species -> total copies
CNV_total <- CNV_correct %>%
  group_by(gene, strain, species) %>%
  summarise(n_copies = sum(n_copies), .groups = "drop")

head(CNV_total)
```


```{r export_fasta}
# ====================
# Export selected hit sequences in FASTA format per strain
# ====================

# Load amino acid sequences from all .faa files
faa_files <- list.files(faa_dir, pattern = "\.faa$", full.names = TRUE)
seq_df <- map_dfr(faa_files, function(f){
  fa <- readAAStringSet(f)
  tibble(header = names(fa), sequence = as.character(fa))
})

# Export sequences for each strain separately
walk2(correct_hits, names(correct_hits), function(d, name) {
  d <- d %>%
    mutate(full_accession = accession) %>%
    left_join(seq_df, by = c("full_accession" = "header")) %>%
    mutate(header = paste0(">", gene, "_", full_accession)) %>%
    filter(!is.na(sequence))
  
  fasta_lines <- map2_chr(d$header, d$sequence, ~paste(.x, .y, sep = "\n"))
  output_path <- file.path(faa_dir, paste0(name, ".faa"))
  writeLines(fasta_lines, output_path)
})
```
